---
title: "Summaries"
date: "2019-03-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyverse)
library(nycflights13)
```

dplyr summarise() function can be used to calculate counts and proportions of logical values: sum(x > 10), mean(y == 0). 

When used with numeric functions, TRUE is converted to 1 and FALSE to 0. 

> This makes sum() and mean() very useful: sum(x) gives the number of TRUEs in x, and mean(x) gives the proportion.

### Exercises

1. Using nycflights13 "flights" dataset, brainstorm at least 5 different ways to assess the typical delay characteristics of a group of flights. 

```{r}
flights
```

```{r}
vars <- c("dep_time", "dep_delay", "arr_time", "arr_delay")
select(flights, dep_time, dep_delay, arr_time, arr_delay)
select(flights, starts_with("dep"), starts_with("arr"))
select(flights, one_of(vars))
select(flights, matches("^dep|^arr"))
select(flights, matches("time$|delay$"), -contains("sched"), -contains("air"))
```

Consider the following scenarios:

  - A flight is 15 minutes early 50% of the time, and 15 minutes late 50% of the time.

```{r}
flights %>%
  group_by(flight) %>%
  summarise(n = n(),
            fifteen_early = mean(arr_delay == -15, na.rm = T),
            fifteen_late = mean(arr_delay == 15, na.rm = T)) %>% 
  filter(fifteen_early == 0.5, fifteen_late == 0.5)
```


  - A flight is always 10 minutes late.

```{r}
flights %>%
  group_by(flight) %>%
  summarise(n = n(),
  ten_always = mean(arr_delay == 10, na.rm = T))%>% 
  filter(ten_always == 1)
```

  - A flight is 30 minutes early 50% of the time, and 30 minutes late 50% of the time.
  
```{r}
flights %>%
  group_by(flight) %>%
  summarise(n = n(),
            thirty_early = mean(arr_delay == -30, na.rm = T),
            thirty_late = mean(arr_delay == 30, na.rm = T)) %>%
  filter(thirty_early == 0.5 & thirty_late == 0.5)
```


  - 99% of the time a flight is on time. 1% of the time it’s 2 hours late.

```{r}
flights %>%
  group_by(flight) %>%
  summarise(n = n(),
            percentage_on_time = mean(arr_delay == 0, na.rm = T),
            twohours = mean(arr_delay > 120, na.rm = T)) %>%
  filter(percentage_on_time == 0.99 & twohours == 0.01)
```


  - Which is more important: arrival delay or departure delay?
Arraival delay.

2. Come up with another approach that will give you the same output as not_cancelled %>% count(dest) and not_cancelled %>% count(tailnum, wt = distance) (without using count()).

```{r}
flights %>% 
  filter(!is.na(dep_delay), !is.na(arr_delay)) %>%
  group_by(tailnum) %>%
  summarize(n = sum(distance))
```


3. Our definition of cancelled flights (is.na(dep_delay) | is.na(arr_delay) ) is slightly suboptimal. Why? Which is the most important column?

There are no flights which arrived but did not depart, so we can just use `!is.na(dep_delay)`.
 We can ensure that a flight is cancelled just wih air_time column.
```{r}
flights %>%
    group_by(departed = !is.na(dep_delay), arrived = !is.na(arr_delay)) %>%
    summarise(n=n())
```


4. Look at the number of cancelled flights per day. Is there a pattern? Is the proportion of cancelled flights related to the average delay?
```{r}
flights %>%
  mutate(dep_date = lubridate::make_datetime(year, month, day)) %>%
  group_by(dep_date) %>%
  summarise(cancelled = sum(is.na(dep_delay)), 
            n = n(),
            mean_dep_delay = mean(dep_delay,na.rm=TRUE),
            mean_arr_delay = mean(arr_delay,na.rm=TRUE)) %>%
    ggplot(aes(x= cancelled/n)) + 
    geom_point(aes(y=mean_dep_delay), colour='blue', alpha=0.5) + 
    geom_point(aes(y=mean_arr_delay), colour='red', alpha=0.5) 
```

5. Which carrier has the worst delays? Challenge: can you disentangle the effects of bad airports vs. bad carriers? Why/why not? (Hint: think about flights %>% group_by(carrier, dest) %>% summarise(n()))

```{r}
flights %>%
    filter(arr_delay > 0) %>%
    group_by(carrier) %>%
    summarise(average_arr_delay = mean(arr_delay, na.rm=TRUE)) %>%
    arrange(desc(average_arr_delay))
```

6. What does the sort argument to count() do. When might you use it?
When you want to sort the cases based on the count.
